---
title: "STAT380 Final Project"
author: Grace Rovinsky, Jiya Patel, Min Liu
format: pdf
---

1. Set the stage–explain the context you're trying to learn something about. MUST ANSWER: Why is this important to you?
Have you ever wondered why a hummingbird needs to eat constantly while a
tortoise can go weeks without food? The answer comes down to metabolism —
the rate at which an organism burns energy to stay alive. Metabolism
affects nearly everything about how an animal lives, from its lifespan to
its diet to how it copes with a changing climate.

As data science students, we got curious about whether the numbers behind
metabolism could actually tell us something deeper — specifically, whether
metabolic rate alone is enough to distinguish between fundamentally
different types of life, like mammals, birds, fish, and insects.

Our research question is: Can we use metabolic rate to classify different
groups of organisms?

This question matters because metabolism is basically life's engine. If we
can predict an organism's group just from how fast it burns energy, that
tells us something profound about how life is organized on Earth — and
could help scientists better understand how different species will respond
as our planet continues to change.

2. Explain the data provenance – Answer the Who, What, When, Where, Why, and How questions for the origin of the data. Include any additional information that would be useful in understanding the context in which the data are embedded.
"Drawing on 20+ published sources, Tori M. Hoehler et al. have compiled a dataset of 10,000+ measurements of metabolic rates of mammals, fish, birds, insects, tree saplings, bacteria, and other living organisms. The dataset includes several types of metabolic rates (primarily basal, field, and maximum), and a mix of individual-organism and species-average measurements."
https://www.pnas.org/doi/10.1073/pnas.2303764120
https://www.pnas.org/doi/10.1073/pnas.2303764120#supplementary-materials
https://en.wikipedia.org/wiki/Basal_metabolic_rate

The data comes from a 2023 study by Tori M. Hoehler and colleagues,
published in the Proceedings of the National Academy of Sciences. They
pulled together over 10,000 metabolic rate measurements from 20+ published
studies and standardized everything to the same unit — watts at 25°C — so
that wildly different organisms like bacteria and mammals could actually
be compared side by side.

The dataset covers a huge range of life including mammals, birds, fish,
insects, bacteria, and tree saplings, and includes information like body
mass, temperature, and type of metabolic rate (basal, field, or maximum).
These variables are particularly important for our analysis because body
mass and metabolic rate are known to vary systematically across organism
groups, making them strong candidates for classification.

This dataset was the perfect fit for our research question — can metabolic
rate be used to classify different groups of organisms — because it
provides standardized, cross-species data at a scale large enough to
train and test machine learning models.

## 3. Exploratory Data Analysis (first draft, please review)

This is a preliminary pass. The goal here is just to get a feel for what we are working with before we lock in the full EDA. Open questions for the team are at the bottom of this section.

### 3.1 Quick look at the dataset

The dataset has about 10,500 rows and 27 columns. The columns we care about most for our research question are `Group`, `Wet Mass (g)`, and `Metabolic Rate (W, at 25C)`. Most rows have all three, so dropping the few rows that are missing any of them does not cost us much. I also drop rows where mass or metabolic rate are zero or negative, since those would break the log scale we use later.

```{r}
library(tidyverse)
library(readxl)

metabolic_data <- read_excel("pnas.2303764120.sd01.xlsx", sheet = "Metabolic_Data")

clean_data <- metabolic_data %>%
  filter(!is.na(`Metabolic Rate (W, at 25C)`),
         !is.na(`Wet Mass (g)`),
         `Wet Mass (g)` > 0,
         `Metabolic Rate (W, at 25C)` > 0,
         !is.na(Group))

# How many measurements per group?
clean_data %>%
  count(Group, sort = TRUE) %>%
  head(10)
```

Fishes have by far the most measurements, followed by mammals and birds. Some groups only have a handful of rows, which is something to keep in mind when we move to modeling, since the model will not have much to learn from those small groups.

### 3.2 First-draft visualization

For a first chart, I plotted body mass against metabolic rate on a log-log scale, because both variables span many orders of magnitude. On a regular scale most points would collapse into the corner. Coloring by group lets us see how each group sits along the trend.

```{r}
top_groups <- clean_data %>%
  count(Group, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Group)

viz_data <- clean_data %>%
  filter(Group %in% top_groups)

ggplot(viz_data, aes(x = `Wet Mass (g)`, y = `Metabolic Rate (W, at 25C)`,
                     color = Group)) +
  geom_point(alpha = 0.4, size = 0.9) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Body mass vs metabolic rate, by group",
       x = "Wet mass (g, log scale)",
       y = "Metabolic rate (W at 25C, log scale)") +
  theme_minimal()
```

Two things stand out. First, there is a clear positive trend across all groups, which matches what we would expect from Kleiber's law about metabolism scaling with body size. Second, the groups do not all sit on top of each other, since bacteria are in the lower left, fishes and insects in the middle, and mammals and birds in the upper right. The overlap is real though, so we will have to see in §4 how well a model can actually pull the groups apart.

### 3.3 Boxplot of metabolic rate by group

```{r}
ggplot(clean_data, aes(x = reorder(Group, log(`Metabolic Rate (W, at 25C)`), median),
                        y = log(`Metabolic Rate (W, at 25C)`), fill = Group)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Log Metabolic Rate by Organism Group",
       x = "Group",
       y = "Log Metabolic Rate (W at 25°C)") +
  theme_minimal()
```

This box plot shows how metabolic rates differ across organism groups.
We log-transformed the metabolic rate because the raw values span many
orders of magnitude, making patterns hard to see. The plot shows clear
differences between groups — mammals and birds tend to have higher
metabolic rates than fish, insects, and bacteria — which suggests that
metabolism does vary systematically across groups.

### Notes for Grace and Jiya

A few open questions before we finalize the EDA:

- Should we keep all 19 groups in the modeling or only the top 5–10 by sample size? The smaller groups will be hard for the model to learn.
- The dataset has a `Type of Metabolic Rate` column with values like basal, field, maximum, etc. Do we want to add a quick chart on this, since mixing types across groups could be a confounder?
- Anything else we want to look at before moving to §4?

4. Use Machine Learning Algorithms
You must use at least one of each: a Regression-based algorithm (SLR/MLR, logistic regression), and a Non-Regression based algorithm (kNN, CART, random forests).
Using an Unsupervised Learning algorithm (clustering, PCA) is highly encouraged, but not required.
Each algorithm must be explained for a general audience.
All results should be explained.
Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.
```{r}
clean_data <- metabolic_data %>%
  filter(!is.na(`Metabolic Rate (W, at 25C)`),
         !is.na(`Wet Mass (g)`),
         !is.na(Group)) %>%
  mutate(log_metabolic_rate = log(`Metabolic Rate (W, at 25C)`),
         log_mass = log(`Wet Mass (g)`),
         Group = as.factor(Group))
```
```{r}
library(caret)
library(class)

## Prepare data
model_data <- clean_data %>%
  select(Group, log_metabolic_rate, log_mass) %>%
  drop_na()

## Train/test split
set.seed(380)
train_index <- createDataPartition(model_data$Group, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

## Run kNN with k=5
train_x <- train_data %>% select(log_metabolic_rate, log_mass)
test_x <- test_data %>% select(log_metabolic_rate, log_mass)
train_y <- train_data$Group

knn_preds <- knn(train_x, test_x, train_y, k = 5)
knn_acc <- mean(knn_preds == test_data$Group)
cat("kNN Accuracy:", round(knn_acc * 100, 2), "%")
```
Our kNN model correctly classified organisms into their groups 72% of
the time using only metabolic rate and body mass. Given that we are
trying to distinguish between many different groups of life — mammals,
birds, fish, insects, bacteria and more — this is a strong result.
It suggests that metabolic rate and body mass together carry real
information about what kind of organism something is. The remaining
28% of misclassifications likely occur between groups that overlap
in metabolic space, such as smaller mammals and larger insects.

5. A discussion of what you've learned from your work about the context and recommendations for future work.

6. Author Contributions – list out who did what in the project.
Jiya - The first boxplot visual, setting the stage, explain data provenance, KNN model
Min - Preliminary EDA (§3.1) and first-draft mass vs metabolic rate visualization (§3.2).

7. References – Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
Hoehler, T. M., et al. (2023). The metabolic rate of the biosphere and its components. Proceedings of the National Academy of Sciences, 120(25). https://www.pnas.org/doi/10.1073/pnas.2303764120

Wikipedia contributors. (2024). Basal metabolic rate. Wikipedia. https://en.wikipedia.org/wiki/Basal_metabolic_rate

8. Code Appendix - Add comments
## Code Appendix
```{r}
## Load all necessary packages
library(readxl)
metabolic_data <- read_excel("pnas.2303764120.sd01.xlsx", sheet = "Metabolic_Data")
```
