---
title: "STAT380 Final Project"
author: Grace Rovinsky, Jiya Patel
format: pdf
---

1. Set the stage–explain the context you’re trying to learn something about. MUST ANSWER: Why is this important to you?
Have you ever wondered why a hummingbird needs to eat constantly while a 
tortoise can go weeks without food? The answer comes down to metabolism — 
the rate at which an organism burns energy to stay alive. Metabolism 
affects nearly everything about how an animal lives, from its lifespan to 
its diet to how it copes with a changing climate.

As data science students, we got curious about whether the numbers behind 
metabolism could actually tell us something deeper — specifically, whether 
metabolic rate alone is enough to distinguish between fundamentally 
different types of life, like mammals, birds, fish, and insects.

Our research question is: Can we use metabolic rate to classify different 
groups of organisms?

This question matters because metabolism is basically life's engine. If we 
can predict an organism's group just from how fast it burns energy, that 
tells us something profound about how life is organized on Earth — and 
could help scientists better understand how different species will respond 
as our planet continues to change.

2. Explain the data provenance – Answer the Who, What, When, Where, Why, and How questions for the origin of the data. Include any additional information that would be useful in understanding the context in which the data are embedded.
"Drawing on 20+ published sources, Tori M. Hoehler et al. have compiled a dataset of 10,000+ measurements of metabolic rates of mammals, fish, birds, insects, tree saplings, bacteria, and other living organisms. The dataset includes several types of metabolic rates (primarily basal, field, and maximum), and a mix of individual-organism and species-average measurements."
https://www.pnas.org/doi/10.1073/pnas.2303764120
https://www.pnas.org/doi/10.1073/pnas.2303764120#supplementary-materials
https://en.wikipedia.org/wiki/Basal_metabolic_rate

The data comes from a 2023 study by Tori M. Hoehler and colleagues, 
published in the Proceedings of the National Academy of Sciences. They 
pulled together over 10,000 metabolic rate measurements from 20+ published 
studies and standardized everything to the same unit — watts at 25°C — so 
that wildly different organisms like bacteria and mammals could actually 
be compared side by side.

The dataset covers a huge range of life including mammals, birds, fish, 
insects, bacteria, and tree saplings, and includes information like body 
mass, temperature, and type of metabolic rate (basal, field, or maximum). 
These variables are particularly important for our analysis because body 
mass and metabolic rate are known to vary systematically across organism 
groups, making them strong candidates for classification.

This dataset was the perfect fit for our research question — can metabolic 
rate be used to classify different groups of organisms — because it 
provides standardized, cross-species data at a scale large enough to 
train and test machine learning models.

3. Exploratory Data Analysis
make 3-4 visualizations
```{r}
library(ggplot2)
library(tidyverse)

clean_data <- metabolic_data %>%
  filter(!is.na(`Metabolic Rate (W, at 25C)`),
         !is.na(Group)) %>%
  mutate(log_metabolic_rate = log(`Metabolic Rate (W, at 25C)`),
         Group = as.factor(Group))

ggplot(clean_data, aes(x = reorder(Group, log_metabolic_rate, median), 
                        y = log_metabolic_rate, fill = Group)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Log Metabolic Rate by Organism Group",
       x = "Group",
       y = "Log Metabolic Rate (W at 25°C)") +
  theme_minimal()

```
This box plot shows how metabolic rates differ across organism groups. 
We log-transformed the metabolic rate because the raw values span many 
orders of magnitude, making patterns hard to see. The plot shows clear 
differences between groups — mammals and birds tend to have higher 
metabolic rates than fish, insects, and bacteria — which suggests that 
metabolism does vary systematically across groups.

4. Use Machine Learning Algorithms
You must use at least one of each: a Regression-based algorithm (SLR/MLR, logistic regression), and a Non-Regression based algorithm (kNN, CART, random forests).
Using an Unsupervised Learning algorithm (clustering, PCA) is highly encouraged, but not required.
Each algorithm must be explained for a general audience.
All results should be explained.
Where appropriate, comparisons between methods should be discussed. This includes highlighting the strengths and weaknesses of the various approaches as well as when each is appropriate.
```{r}
clean_data <- metabolic_data %>%
  filter(!is.na(`Metabolic Rate (W, at 25C)`),
         !is.na(`Wet Mass (g)`),
         !is.na(Group)) %>%
  mutate(log_metabolic_rate = log(`Metabolic Rate (W, at 25C)`),
         log_mass = log(`Wet Mass (g)`),
         Group = as.factor(Group))
```
```{r}
library(caret)
library(class)

## Prepare data
model_data <- clean_data %>%
  select(Group, log_metabolic_rate, log_mass) %>%
  drop_na()

## Train/test split
set.seed(380)
train_index <- createDataPartition(model_data$Group, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

## Run kNN with k=5
train_x <- train_data %>% select(log_metabolic_rate, log_mass)
test_x <- test_data %>% select(log_metabolic_rate, log_mass)
train_y <- train_data$Group

knn_preds <- knn(train_x, test_x, train_y, k = 5)
knn_acc <- mean(knn_preds == test_data$Group)
cat("kNN Accuracy:", round(knn_acc * 100, 2), "%")
```
Our kNN model correctly classified organisms into their groups 72% of 
the time using only metabolic rate and body mass. Given that we are 
trying to distinguish between many different groups of life — mammals, 
birds, fish, insects, bacteria and more — this is a strong result. 
It suggests that metabolic rate and body mass together carry real 
information about what kind of organism something is. The remaining 
28% of misclassifications likely occur between groups that overlap 
in metabolic space, such as smaller mammals and larger insects.

5. A discussion of what you’ve learned from your work about the context and recommendations for future work.

6. Author Contributions – list out who did what in the project.
Jiya - The first boxplot visual, setting the stage, explain data provenance, KNN model

7. References – Citation style is your choice, but all sources should be documented (both in text and in the References section). This includes where you got your data.
Hoehler, T. M., et al. (2023). The metabolic rate of the biosphere and its components. Proceedings of the National Academy of Sciences, 120(25). https://www.pnas.org/doi/10.1073/pnas.2303764120

Wikipedia contributors. (2024). Basal metabolic rate. Wikipedia. https://en.wikipedia.org/wiki/Basal_metabolic_rate

8. Code Appendix - Add comments
## Code Appendix
```{r}
## Load all necessary packages
library(readxl)
metabolic_data <- read_excel("~/Downloads/pnas.2303764120.sd01.xlsx", sheet = "Metabolic_Data")
```
